# Examining these procedures with the 2022 CES

How would I implement these procedures with a common political science dataset? This section demonstrates how to do this with the common content from the 2022 Cooperative Election Study (CES). While I am unable to compare the ability for these procedures to reduce bias as I do not have complete data, the next section provides computational benchmarks for each of these procedures on a common dataset in political science as well as estimates the Z-score difference between these procedures to demonstrate whether resulting differences between these procedures' performance has any meaningful difference that the practitioner should be concerned with.

I first clean the 2020 CES by removing columns reporting the respondents' case number, whether they took the CCES and the post-election questionnaire, how long respondents spent on any given page, non-numeric columns, and questions in the post-election questionnaire that were also asked in the pre-election questionnaire.

```{r}
#| label: load-cces

if (params$reproduce == TRUE) {
  source("2022_ces_cleaning.R")
} else {
  load(
    file = "../data/2020_cces/cleaned.RData"
  )
}
```

Once the cleaning was complete, I generate five imputed datasets with each procedure used in the simulation study. While performing the imputation, I use the `tictoc` [@tictoc] package to record the amount of time each imputation task took to complete. Once the imputation was completed, 

```{r}
#| label: impute-ces

if (params$reproduce == TRUE) {
  tic.clearlog() # clear log for tic
  #* mean
  tic("Interpolate")
  df_mean <- mice::mice(
    df_ces_clean
    , m = 5
    , meth = "mean"
    , printFlag = FALSE
  )
  toc(
    log = TRUE
    , quiet = TRUE
  )
  #* Amelia
  tic("Amelia")
  df_amelia <- Amelia::amelia(
    df_ces_clean
    , m = 5
    , p2s = 0
  )
  toc(
    log = TRUE
    , quiet = TRUE
  )
  #* LMICE
  tic("L-MICE")
  df_lmice <- mice::mice(
    df_ces_clean
    , m = 5
    , meth = "norm"
    , printFlag = FALSE
  )
  toc(
    log = TRUE
    , quiet = TRUE
  )
  #* RFMICE
  tic("RF-MICE")
  df_rfmice <- mice::mice(
    df_ces_clean
    , m = 5
    , meth = "rf"
    , printFlag = FALSE
  )
  toc(
    log = TRUE
    , quiet = TRUE
  )
  # miceRanger
  tic("miceRanger")
  df_miceRanger <- miceRanger::miceRanger(
    df_ces_clean
    , m = 5
    , verbose =  FALSE
  )
}
```
